---
title: "Not-Project1"
author: "Kirsten Bauck"
date: "2025-02-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(knitr)
library(gridExtra)
library(corrplot)
# Decision Tree
library(tree)
library(rpart)
library(partykit)
# Random Forest
library(caret)
library(randomForest)
# AdaBoost
library(adabag)
```


# Introduction
This is the kaggle link: https://www.kaggle.com/datasets/fedesoriano/stellar-classification-dataset-sdss17
• Why did you choose this sample?
• What did you hope to learn?
• What is important/interesting about the data?
• How confident do you feel in the source of the data?


# Data Cleaning
• What was necessary to get the data properly cleaned?
• What challenges did you encounter while cleaning?
• A short description of key variables in the data set.
• A data table presenting an abbreviated tidydata set.

```{r Data-Cleaning, echo=False}
set.seed(578)
stars = read.csv("star_classification.csv")

# Remove extreme outliers
outliers = boxplot.stats(stars$u[stars$class == "STAR"])$out
removed = stars[stars$u %in% outliers & stars$class == "STAR", ][9, ]
stars = stars[-79544, ]

# Checking for rows NA values
rows_with_na = stars[!complete.cases(stars), ]
# Checking for duplicate values
rows_duplicated = stars[duplicated(stars), ]

```

# Data Visualizations
• Figures/graphs are to be presented on their own page.
• Each figure should come a detailed description/caption.
• These should come after the intro and cleaning paragraphs.
• Descriptions provided additional to the caption should be included in normal paragraph text.
• Focus on what the visualization says about achieving an outcome, like prediction of a response, or alignment of groupings (such as prediction of a categorical
response)

# Conclusions
• If given more time, what other statistical or exploratory analysis might you have done?
• If interested, attempt a statistical or additional exploratory analysis.
• What did you learn if you used this data in a previous project?
• What challenges in analysis would you foresee

# Link to GitHub Repository
Something like: devtools::install_github('DrBuscaglia/F23DemoPackage')